A Survey of Quantitative Descriptions of Molecular Structure

### 2012-04-09 paper.tex
Started on polymer descriptors
Added some more text on qm descs

Geometric descriptors require a 3D conformation as input and therefore
are not as fast as topological descriptors which do not require the 3D
coordinate generation step. Nearly all geometric descriptors work on a
single conformation and for cases where multiple conformation must be
considered, an averaging procedure is usually employed.

A number of geometric descriptors characterize the molecular
surface. The simplest such example is the surface area descriptor (and
the corresponding volume descriptor). A simple extension of such a
descriptor is to characterize distributions of a physicochemical
property over the molecular surface such as the Charged Partial
Surface Area descriptors \cite{Stanton:1990aa} that characterize the
partial charge distribution over the molecular surfce. Note that
surface descriptor calculations based on analytical
\cite{Connolly:1983aa} or tesselation algorithms
\cite{Eisenhaber:1995qf} can be slow and that parametrized methods
such as the VSA descriptors \cite{Labute:2008aa} that are based on
precalculated surface area values derived from a list of functional
groups. A comparison of surface area values generated by the two
approaches for XXX molecules indicates a high degree of
correlation.

While molecular surfaces are useful, molecular shape is a key factor
in ligand-receptor interactions. A variety of descriptors have been
devised to characterize molecular shape. The use of overlapping atom
centered Gaussians to represent molecular shape \cite{Grant:1995aa}
represents a significant computational advantage over the traditional
analytical approach of intersecting hard spheres, supporting extremely
rapid shape comparisons \cite{Grant:1999vn}. An alternative approach
described by Ritchie et al \cite{Ritchie:1999kx} employed a spherical
harmonic based representation of surfaces, that in contrast to the
Gaussian approach allows one to generate characterizations of the
surface and hence the shape at varying resolutions. Another approach
to multi-resolution shape representations is the use of $\alpha$
shapes \cite{Edelsbrunner:1994aa}, which represent a parametrization
of the convex hull derived from the molecular coordinates. This
approach has been employed to characterize shape and volume
\cite{Wilson:2009ys} though it is more computationally intensive than
the methods mentioned previously. In addition to functional
representations of molecular shape, a number of simpler formulations
have been defined. For example, the ``shape signature'' approach
described by Zauhar et al \cite{Zauhar:2003fk} is a novel application
of a ray tracing method that generates a description of the molecular
shape. Ballester et al \cite{Ballester:2007aa} described the Ultrafast
Shape Recognition (USR) approach that characterized molecular shape in
terms of the distribution of distances between the atoms and a set of
pre-defined points. This procedure characterizes the shape in the form
of a 12-D real valued fingerprint. As a result, the calculation
procedure is extremely fast and once calculated shape similarities can
be evaluated rapidly via a simple distance calculation.

Finally, quantum mechanical calculations can also generated
descriptors that can be applied to problems such as QSAR models
(though of course the time complexity of QM derived descriptors do not
make them amenable for large scale virtual screening applications).
\textbf{QM descriptors}

@@

Polymeric systems represent a challenge for descriptor development. A
number of approaches have been described that includes both
deterministic, geometry derived descriptors \cite{Edvinsson:2003bh} as
well as stochastic methods \cite{Gonzales-Diaz:2003ly}.

>>>>

Geometric descriptors require a 3D conformation as input and therefore
are not as fast as topological descriptors which do not require the 3D
coordinate generation step. Nearly all geometric descriptors work on a
single conformation and for cases where multiple conformation must be
considered, an averaging procedure is usually employed. Examples of
these descriptors include the gravitational index [REF] and moment of
inertia descriptors. However, geometric descriptors encompass many
aspects of a molecular structure. For example, a number of geometric
descriptors characterize the molecular surface. The simplest such
example is the surface area descriptor (and the corresponding volume
descriptor). A simple extension of such a descriptor is to
characterize distributions of a physicochemical property over the
molecular surface such as the Charged Partial Surface Area descriptors
\cite{Stanton:1990aa} that characterize the partial charge
distribution over the molecular surfce. Note that surface descriptor
calculations based on analytical \cite{Connolly:1983aa} or tesselation
algorithms \cite{Eisenhaber:1995qf} can be slow and that parametrized
methods such as the VSA descriptors \cite{Labute:2008aa} that are
based on precalculated surface area values derived from a list of
functional groups. A comparison of surface area values generated by
the the tesselatin method and the VSA method (as implemented in the
CDK) for XXX molecules indicates a high degree of correlation.

Molecular shape is a key factor in ligand-receptor interactions and is
an important approach to virtual screening. A variety of descriptors
have been devised to characterize molecular shape. The use of
overlapping atom centered Gaussians to represent molecular shape
\cite{Grant:1995aa} represents a significant computational advantage
over the traditional analytical approach of intersecting hard spheres,
supporting extremely rapid shape comparisons \cite{Grant:1999vn}. An
alternative approach described by Ritchie et al \cite{Ritchie:1999kx}
employed a spherical harmonic based representation of surfaces, that
in contrast to the Gaussian approach allows one to generate
characterizations of the surface and hence the shape at varying
resolutions. Another approach to multi-resolution shape
representations is the use of $\alpha$ shapes
\cite{Edelsbrunner:1994aa}, which represent a parametrization of the
convex hull derived from the molecular coordinates. This approach has
been employed to characterize shape and volume \cite{Wilson:2009ys}
though it is more computationally intensive than the methods mentioned
previously. In addition to functional representations of molecular
shape, a number of simpler formulations have been defined. For
example, the ``shape signature'' approach described by Zauhar et al
\cite{Zauhar:2003fk} is a novel application of a ray tracing method
that generates a description of the molecular shape. Ballester et al
\cite{Ballester:2007aa} described the Ultrafast Shape Recognition
(USR) approach that characterized molecular shape in terms of the
distribution of distances between the atoms and a set of pre-defined
points. This procedure characterizes the shape in the form of a 12-D
real valued fingerprint. As a result, the calculation procedure is
rapid and once calculated shape similarities can be evaluated rapidly
via a simple distance calculation.

Finally, quantum mechanical calculations can also generated
descriptors that can be applied to problems such as QSAR models
(though of course the time complexity of QM derived descriptors do not
make them amenable for large scale virtual screening
applications). These descriptors can be evaluated at various levels of
theory. For example, partial charges, HOMO and LUMO energies are
commonly evaluated using semi-empirical methods, while Eroglu and
T\"{u}rkmen \cite{Eroglu:2007uq} described DFT-derived descriptors
such as electronegativity, hardness and ionization potential (though
Puzyn et al question the need for such a high level of theory
\cite{Puzyn:2008fk}).

We next consider descriptors that are designed for specific classes of
molecular systems.

@@

Polymeric systems represent a challenge for descriptor
development. There have been reports that employed tradtional
small-molecule descriptors \cite{Mattioni:2002dq,Katritzky:1998cr}
that consider the polymer in terms of the individual monomers. A
number of approaches have been described that includes both
deterministic, geometry derived descriptors \cite{Edvinsson:2003bh} as
well as stochastic methods \cite{Gonzales-Diaz:2003ly}.
